---
title: "MHCIIDQB Phylogenetic Tree of the Irrawaddy dolphin in Relation to Cetaceans"
author: "Sett Aung"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
---

```{r load-packages, message=FALSE}
library(tidyverse)
library(openintro)
library(ggtree)
library(ape)
```


## Introduction
Species become endangered because of habitat loss and genetic variation. One of the species at the risk of extinction are the Irrawaddy dolphin (Orcaella brevirostris) listed as endangered by the International Union for Conservation of Nature (IUCN). (Minton, et al. 2017). The Irrawaddy dolphin are small cetacean species that inhabit throughout South-East Asia rivers, estuarine, and coastal environment distributing from North-East India, Myanmar, Thailand, Cambodia, Laos, Vietnam, Indonesia to the east of the Philippines. In coastal habitat, are often associated with freshwater discharges, and estuaries (Jackson Ricketts, et al. 2018). The freshwater subpopulation occurs in three large rivers which includes the Ayeyarwaddy (Irrawaddy) River in Myanmar, Mahakam River in Indonesia, and Mekong River in Cambodia and Southern Laos. This species is also found as a lacustrine subpopulation in lakes such as Chilika Lake in India, Songkhla Lagoon in Thailand, and Malampaya Sound in the Philippines (Beasley et al. 2002, Dai et al. 2021). The riverine subpopulation are considered critically endangered by IUCN and their long-term survival has become a conservative concern due to their fragmented distribution and threats from human activities such as electrofishing and entanglement in gillnets (Gilbert et al. 2009, Brownell et al. 2019), habitat loss from dam constructions in riverine populations (Brownell et al. 2017 ), contamination from pollutants (such as oil, pesticides, and industrial wastes) in the estuarine population (Schnitzler et al. 2021, Murphy et al. 2007), and vessel strikes in some areas (Thomas et al. 2019). With restricted dispersal, and possible population bottlenecks, some isolated subpopulations will be subjected to inbreeding resulting in limited gene flow (Frankham, 1995) which could lead to loss of genetic variation (Hamner, et al. 2012).

The major histocompatibility complex (MHC) genes have been widely studied to examine genetic variation and determine population health (Sommer, 2005). MHC class II genes are highly polymorphic, especially the peptide binding region (PBR) that is involved with recognition and binding with pathogen antigens that initiate T-lymphocyte immune response (Santillan et al. 2016; Caballero et al. 2010). Past studies have hypothesized that MHC polymorphism could be maintained by over-dominant selection or a balancing selection that favors heterozygosity or a pathogen mediated selection (Sommer, 2005; Hughes, 1995). It has not been clear which hypothesis supports the polymorphism of MHC.  

Previous research by Hayashi et al. 2003, has found only one version of DQB gene (and only one DQB allele) because of the sample size (n=1). My goal to study genetic variation at the DQB locus in Irrawaddy dolphins across the South-East Asia distribution range is to determine the Irrawaddy dolphin population health for conservation and provide more information about the mechanism in maintaining variability of MHC genes.  Therefore, I propose to test the following goals:

## Experimentnal Design

With thanks to Dr. Michael McGowen, the marine mammal curator from Smithsonian National Museum of Natural History, we obtained Irrawaddy dolphin teeth and bone samples from recently stranded individuals from the Department of Marine and Coastal Resources (DMCR) in Klaeng, Rayong, Thailand (n=11) and the MareCet Research Organization (MareCet) in Selangor, Malaysia (n=2). We also obtained tissue sample from various sources including the Southwest Fisheries Science Center’s (SWFSC) Marine Mammal and Turtle Molecular Research Tissue Sample Collection (MMASTR) in La Jolla, CA, USA (n=29), the DMCR (n=11), and the MareCet (n=6).  Since the species is under risk of extinction, we obtained historic samples originally collected between 1875-1972 which are osteological samples (n=16) and bone samples from Orcaella heinsohni (n=2) stored in eight museum collections in the United States and Europe. Unfortunately, we were not able to obtain modern samples in regions such as Sumatra (Indonesia), Ayeyarwaddy River (Myanmar), Chilika Lake (India), and the northern coast of Borneo.

All 77 samples were DNA extracted, Illumina NGS sequenced, and will be analyzed to reveal potential additional DQB alleles. From this experiment, we can identify additional DQB alleles that may be useful to recover the populations. 

## Aim 2 Construct a phylogenetic tree to compare the DQB alleles from this study with the DQB alleles from cetaceans. 

Rationale: Mechanisms that maintain MHC genetic diversity remain an open question in biology. There have been theories hypothesizing the maintenance of high variability of the MHC genes. This aim will be done to enrich the understanding of the evolution and relationship of the DQB alleles among other cetaceans. 

Null hypothesis: There is no relationship between the MHCIIDQB Alleles of Irrawaddy dolphins with some of the cetaceans.
Alternative hypothesis: There is a relationship between the MHCIIDQB Alleles of Irrawaddy dolphins with some of the cetaceans.

Methods: To analyze the relationship of the DQB alleles with few other cetaceans, we will compile Cetacean MHCII DQB sequences from GenBank and align them to the DQB alleles from Irrawaddy dolphin and input the phylogenetic tree from IQ Tool-CIPRES using R to make the tree better. The data from this approach will help us visualize which group of cetaceans has high similarity with the DQB alleles of Irrawaddy dolphins. This information can clarify more about the maintenance of high variability of MHC genes over time. 

## Using ggtree
```{r}
tree <- read.tree("../Data/output.bionj")
```

```{r}
# Ladderizing the tree (Neighborjoining tree)
tree <- ladderize(tree, right = F)
```

```{r}
# Importing NJ tree from CIPRES using IQ Tool
tree_1 <- ggtree(tree, branch.length='none', layout='circular') +
  geom_tiplab(size = 1) +
  ggtitle(
"Neighbor-joining tree - Comparing DQB Alleles of Irrawaddy 
                dolphins to DQB Alleles of cetaceans ")

tree_1
```

## Using ggtree
```{r}
# reading contree output from CIPRES
tree2 <- read.tree("../Data/output.contree")
```

```{r}
# Ladderizing
tree2 <- ladderize(tree2, right = F)
```

```{r}
# Importing contree from CIPRES using IQ Tool
group = c("Obre_Allele1", "Obre_Allele2", "Obre_Allele3", "Obre_Allele4", "AB164223_Obre")
tree_2 <- ggtree(tree2, branch.length='none', layout='circular') +
  geom_tiplab(size = 1) +
  ggtitle("Bootstrap tree - Comparing DQB Alleles of Irrawaddy dolphins to 
                       DQB Alleles of cetaceans") +
  geom_treescale() 
  
tree_2
```

## Using ggtree
```{r}
# reading ML treefile output from CIPRES
tree3 <- read.tree("../Data/output.treefile")
```

```{r}
# Ladderizing
tree3 <- ladderize(tree3, right = F)
```

```{r}
# Importing ML tree from CIPRES using IQ Tool
tree_3 <- ggtree(tree3, branch.length='none', layout='circular') +
  geom_tiplab(size = 1) +
  ggtitle("Maximum Likelihood tree - Comparing DQB Alleles of Irrawaddy 
                    dolphins to DQB Alleles of cetaceans") +
  geom_treescale() 


tree_3
```

## Using ape package
```{r}
# Load the 'ape' package 
library(ape)
tree2 <- read.tree("../Data/output.contree")
# Ladderizing
tree2 <- ladderize(tree2, right = FALSE)


# Define the labels you want to highlight
highlight_labels <- c("Obre_Allele1", "Obre_Allele2", "Obre_Allele3", "Obre_Allele4", "AB164223_Obre")

# Plot the tree with highlighted labels
plot(tree, tip.color = ifelse(tree$tip.label %in% highlight_labels, "red", "black"), cex = 0.3, 
     type = 'fan')

```
```{r}

```

## References

Bezuijen, M.R., Zanre, R., Goichot, M. 2007. The Don Sahong Dam and the Irrawaddy dolphin. WWF Greater Mekong Programme, Vientiane

Beasley, I., Chooruk, S., and Piwpong, N. 2002. THE STATUS OF THE IRRAWADDY DOLPHIN, ORCAELLA BREVIROSTRIS, IN SONGKHLA LAKE, SOUTHERN THAILAND. THE RAFFLES BULLETIN OF ZOOLOGY: Supplement No. 10: 75-83 

Brownell, Jr. R. L., Reeves, R. R., Read A. J., Smith, B. D., Thomas, P. O., Ralls, K., Amano, M., Berggren, P., Chit, A. M., Collins, T., Currey, R., Dolar, M. L. L., Genov, T., Hobbs, R. C., Kreb, D., Marsh, H., Mei Z., Perrin, W. F., Phay, S. ...Wang, J. Y. 2019. Bycatch in gillnet fisheries threatens Critically Endangered small cetaceans and other aquatic megafauna. Endangered Species Reserach 40, 285-296. https://doi.org/10.3354/esr00994 

Brownell Jr. R.L., Reeves RR, Thomas PO, Smith BD, Ryan GE. Dams threaten rare Mekong dolphins. Science. 2017 Feb 24;355(6327):805. doi: 10.1126/science.aam6406. Epub 2017 Feb 23. PMID: 28232545.

Bolger AM, Lohse M, Usadel B. 2014 Trimmomatic: a flexible trimmer for Illumina sequence data. Bioinformatics 30, 2114-2120. (doi: 10.1093/bioinformatics/btu170)

Budi, T., Piyapattanakorn, S., Kreb, D., Yuda, P., Ninwat, S., Hardwises, P., Prachamkhai, P., Senanan, W., Thongsukdee, S., Phavaphutanon, J., and Klinsawat, W. 2022. Mitogenomes provide insight into complex evolutionary history of freshwater and coastal Irrawaddy dolphin Orcaella brevirostris Gray, 1966) in Thailand and Indonesia. Agriculture and Natural Resources 56, 591-606. https://doi.org/10.34044/j.anres.2022.56.3.15 

Caballero, S., Heimeier, D., Trujillo, F., Vianna, J.A., B-Garrido., Montiel, M.J., Pedereros, S. V., 
Marmontel, M., De. O. Santos, M.C., Santos, F.R., and Baker, C.S. 2010. Initial description of Major Histocompatibility Complex variation at two class II loci (DQA-DQB) in Sotalia fluviatilis and Sotalia guianensis. LAJAM 8(1-2): 81-95. e-ISSN 2236-1057

Cerretero, S.A., Kapli, P., and Yang, Z. 2023. Beginner’s Guide on the Use of PAML to Detect Positive Selection. Mol. Biol. Evol. 40(4):msad041. https://doi.org/10.1093/molbev/msad041 

Dabney J & Meyer M. 2019. Extraction of highly degraded DNA from ancient bones and teeth. In B. Shapiro, A. Barlow, P. Heintzman, M. Hofreiter, J. Paijmans, & A. Soares (Eds.), Ancient DNA, Methods in Molecular Biology vol. 1963, pp. 25-29. Humana Press. https://doi.org/10.1007/978-1-4939-9176-1_4

Dai, Y., Chantra, R., Kittiwattanawong, K., Zhao, L., Sakornwimon, W., Aierken. R., Wu, F., and Wang, X. 2021. Genetic structure of the endangered Irrawaddy dolphin (Orcaella brevirostris) in the Gulf of Thailand. Genetics and Molecular Biology 44, e20200365. https://doi.org/10.1590/1678-4685-GMB-2020-0365 

Frankham, R. 1995. Conservation Genetics. Annu. Rev. Genet. 29(1):305–327
IUCN–SSC Cetacean Specialist Group, “Mekong River Irrawaddy dolphins.”. 2021. https://iucn-csg.org/mekong-dolphins/

Hamner, R.M., Pichler, F.B., Heimeier, D., Constantine, R., and Scott, B. C. 2012.  Genetic differentiation and limited gene flow among fragmented populations of New Zealand endemic Hector’s and Maui’s dolphins. Conserv Genet 13, 987–1002. 
https://doi.org/10.1007/s10592-012-0347-9

Hayashi, K., Nishida, S., Yoshida, H., Goto, M., Pastene, L.A., and Koike, H. 2003. Sequence variation of the DQB allele in the cetacean MHC. Mammal Study 28: 89–96. DOI: 10.3106/mammalstudy.28.89 

Hughes, A.L., and Hughes, M.K. 1995. Natural Selection of the peptide binding regions of major histocompatibility complex molecules. Immunogenetics 42: 233-243. 

Jackson-Ricketts, J., Ruiz-Cooley, R. I., Junchompoo, C., Thongsukdee, S., Intongkham, A., Ninwat, S., Kittiwattanawong, K., Hines, E. M., & Costa, D. P. (2019). Ontogenetic variation in diet and habitat of Irrawaddy dolphins (Orcaella brevirostris) in the Gulf of Thailand and Andaman Sea. Marine Mammal Science 35, 492-521. https://doi.org/10.111/mms.12547 

Miller, M.A., Pfeiffer, W., and Schwartz, T. 2011. The CIPRES science gateway: a community resource for phylogenetic analyses. Association for Computing Machinery, New York, NY, USA, Article 41, 1-8. https://doi.org/10.1145/2016741.2016785

Minton G, Smith BD, Braulik GT, Kreb D, Sutaria D and Reeves R. 2017. Orcaella brevirostris (errata version published in 2018), The IUCN Red List of Threatened Species, https:// www.iucnredlist.org/species/15419/12379085

Moreno-Santillán, D.D., Lacey, E.A., Gendron,, D., and Ortega, J. 2016. Genetic Variation at Exon of the MHC Class II DQB Locus in Blue Whale (Balaenoptera musculus) from the Gulf of California. PLoS ONE 11(1): e0141296. doi:10.1371/journal. pone.0141296

Murphy, P.T., Irvine, K.N., Sampson, M., Guo, J., and Parr T.  2007. Mercury contamination along the Mekong River, Cambodia. Asian Journal of Water, Environment and Pollution. 6(1): 1–9

Nei, M., and Gojobori, T. 1986. Simple methods for estimating the numbers of synonymous and nonsynonymous nucleotide substitutions. Molecular Biology and Evolution 3:418-426.

Schnitzler, J.G., Das, K., Wohlsein, P., Kuiken, T., Ludwig, A., Lieckfeldt, D., Phan, C., Phay, S., and Siebert, S. 2021. The Irrawaddy dolphin (Orcaella brevirostris) from the Mekong river Cambodia: Preliminary health and toxicological investigations. Aquatic Toxicol., 234. https://doi.org/10.1016/j.aquatox.2021.105812

Sommer, S. 2005. The importance of immune gene variability (MHC) in evolutionary ecology and conservation. Front Zool. doi: 10.1186/1742-9994-2-16 PMID: 16242022

Thomas, P. O., Gulland, F. M. D., Reeves, R. R., Kreb, D., Ding, W., Smith, B., Malik, M. I., Ryan, G. E., & Phay, S. 2019. Electrofishing as a potential threat to freshwater cetaceans. Endangered Species Research 39, 207-220. https://doi.org/10.3354/esr00962

Tamura, K., Stecher, G., and Kumar, S. 2021. MEGA 11: Molecular Evolutionary Genetics Analysis Version 11. Molecular Biology and Evolution https://doi.org/10.1093/molbev/msab120.

Stecher, G., Tamura K., and Kumar, S. 2020. Molecular Evolutionary Genetics Analysis (MEGA) for macOS. Molecular Biology and Evolution 37:1237-1239.








